---
title: "USFE Year Month Averages"
author: "Erika W"
date: "6/28/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
```

```{r}
rm(list=ls())
library(ggplot2) # For plotting with ggplot
library(lubridate) # For working with dates in datasheets
library(httr)
library(tidyverse)
library(gh) # for loading from private repo
library(gitcreds)# for loading from private repo
library(dplyr)
```

## Load Data

Start with the final merged CEDEN-SURF dataset that has consistent units

```{r}

### Download CEDEN/SURF data from IETC private repository
tmp <- tempfile()

CS_Limited <- gh("https://raw.githubusercontent.com/WWU-IETC-R-Collab/CEDENSURF-mod/30YRS_OrigRR/Data/Output/CEDENSURF_Limited_FixedUnits.csv",
                 .token = gh_token(),
                 .destfile = tmp)

Limited <- read_csv(tmp)

names(Limited)[names(Limited) == "Subregion"] <- "RiskRegion"

min(Limited$FullDate)
```

## Summarize data by Analyte, Region, Month & Year

The goal is to create a linear set of data with either an average value or NA for every month of every year in the sample time frame. 
We can accomplish this in several steps:

### 1. Create a complete date:month sequence for the data timeframe

```{r}
# Create empty matrix with correct number of rows
  scaffold<- data.frame(matrix(ncol = 1, nrow = 300))
  
  # Fill with complete series of months
  scaffold$Date<- seq(as.Date("1995/1/1"), as.Date("2019/12/31"), "months")
  
  # Simplify to month:date
  Dates <- tidyr::separate(scaffold, Date, c('Year', 'Month', 'Day')) %>%
    select(Year, Month)
  
  head(Dates)
  
  # Create text month column

  MonthTable<- data.frame(Month = c("01", "02", "03", "04", "05", "06", "07", 
                                    "08", "09", "10", "11", "12"),
                          MonthT = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", 
                                     "Aug", "Sep", "Oct", "Nov", "Dec")) %>%
    mutate(Season = case_when(Month %in% c("11", "12", "01", "02", "03", "04") ~ 'Wet', 
                              Month %in% c("05", "06", "07", "08", "09", "10") ~ 'Dry'))
  
  Dates<- merge(Dates, MonthTable, by="Month", all.x = T)

  head(Dates)
  
  #Clean environment for next steps
  rm(scaffold, tmp, CS_Limited, MonthTable)
```

###2. Summarize original data

```{r}
#Split date into Year, Month, Day in separate columns so we can analyze by month 

Limited$FullDate <- Limited$Date # Duplicate date column

Limited <- tidyr::separate(Limited, Date, c('Year', 'Month', 'Day')) # Split original date column by Year, Month, and Day

# summarize
Summary <- Limited %>% 
  filter(Analyte %in% c("bifenthrin", "chlorpyrifos", "diazinon","triadimefon", "myclobutanil",            "paraquat_dichloride", "malathion", 
                "mercury", "copper")) %>%
  group_by(RiskRegion, Year, Month, Analyte) %>%
  summarize(n = n(),
            mean = mean(Result),
            sd = sd(Result),
            min = min(Result),
            max = max(Result))
```

### 3. Merge complete date range with each Analyte:Subregion subset

```{r}

# Create vectors to loop over

Region<- unique(Limited$RiskRegion)

SelectList <- c("bifenthrin", "chlorpyrifos", "diazinon",
                "paraquat_dichloride", "malathion", 
                "mercury", "copper")


# Create empty list to store intermediate / iterated output

Result <- list()

# Double Loop (Let's see how this goes!)


for (i in 1:length(SelectList)){
  A_sub <- Summary %>% filter(Analyte == SelectList[i]) # subset records for one Analyte
  
  for(z in 1:length(Region)){
    B_sub <- A_sub %>% 
      filter(RiskRegion == Region[z]) # subset records for one RR
    
    M <- merge(Dates, B_sub, by=c("Year", "Month"), all = TRUE) ### Merge so that retains date rows that have no matching data in sampling records
    M$RiskRegion <- Region[z] #fill blank rows with appropriate risk region
    M$Analyte <- SelectList[i]
    M$FullDate<- paste(M$Year, M$Month)
    Result[[z]]<- M
  }
  
  M<- do.call(rbind,Result) # recombine all result df 
  N<- M %>% arrange(Year, Month)
  Analyte <- SelectList[i]
  write.csv(N, file= paste("Output/",Analyte[1],".csv"), row.names = F)
}


```

This produces a table with n, mean, sd, etc for every analyte of interest. 

## For Wayne

I understand the request was to produce the data in a similar format as required for Brian's spreadsheets, which required data from 1995 - 2015 in wide format. This requires just a slightly different approach;

```{r}

# Restrict to desired date range
Summary <- Summary %>% filter(between(Year,1995,2015))
     
# Create vectors to loop over

Region<- unique(Limited$RiskRegion)

SelectList <- c("bifenthrin", "chlorpyrifos", "diazinon","triadimefon", "myclobutanil",            "paraquat_dichloride", "malathion", 
                "mercury", "copper")


# Create empty list to store intermediate / iterated output

Result <- list()
  
##  Loop!

  for (i in 1:length(SelectList)){
    A_sub <- Summary %>% filter(Analyte == SelectList[i]) # subset records for one Analyte
    
    for(z in 1:length(Region)){
      B_sub <- A_sub %>% 
        filter(RiskRegion == Region[z]) # subset records for one RR
      
      M <- merge(Dates, B_sub, by=c("Year", "Month"), all = TRUE) ### Merge so that retains date rows that have no matching data in sampling records
      M$RiskRegion <- Region[z] #fill blank rows with appropriate risk region
      M$Analyte <- SelectList[i]
      M$FullDate<- paste(M$Year, M$Month)
      Result[[z]]<- M
    }
    
    M<- do.call(rbind,Result) # recombine all result df 
    N<- M %>% arrange(Year, Month) %>% 
      select(FullDate, RiskRegion, Analyte, mean) %>%
      rename(Result = mean) %>%
      group_by(FullDate, RiskRegion, Analyte)%>%
      pivot_wider(names_from = FullDate,
                  names_repair = "check_unique",
                  values_from = Result)
    
    Analyte <- SelectList[i]
    
    write.csv(N, file=paste("Output/",Analyte[1],"w",".csv"), row.names = F)
  }
```

